---
title: 'Visualizing the Fish Population'
author:
  - Matt Wilson
  - Mark Newman
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
output:
  rmarkdown::html_vignette
bibliography: bib.bib
vignette: >
  %\VignetteIndexEntry{Visualizing the Fish Population}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

When viewing any geo-spacial data, we need to set up a region of interest.

```{r}
state_name <- 'Pennsylvania'
county_name <- 'York'
```

Get the map information for all the counties in `state_name`, then filter that to the `county_name` we want to focus on.
We focus on a single county because otherwise there will be too many small watersheds.

```{r}
library(dplyr)
library(tigris)

map <-
  counties(state = 'Pennsylvania') %>%
  filter(NAME == county_name)
```

Lets look at the region we have so far.
`ggplot2` paints in layers, so this is a very useful exercise.

Hint: For those familiar with `ggplot2`, but not geo-spacial, an `aes()` is implicitly constructed in `geom_sf()` by searching `map` for a polygon type column.
This means that adding `aes(geometry = geometry)` was redundant.

```{r}
library(ggplot2)

ggplot(map, aes(geometry = geometry)) +
  theme_bw() + 
  geom_sf()
```

Lets get all the watersheds in and around our region of interest.
`FedData` provides clean access to this information.

**Hint**: `FedData`'s `get_wbd()` and `` functions can be very slow.
If you have any interest in looking at the data more than once, make sure to use `extraction.dir`.

**Hint**: When requesting a single boundary, `get_wbd()` has an interesting re-run functionality.
On the first call it returns a `data.frame()` of the information we want.
On the second call it returns a `list()` of `data.frame()`.
To provide a consistent experience we need some hackery. 

```{r}
library(FedData)
library(rappdirs)

extraction.dir <- user_cache_dir("FedData")
if(!dir.exists(extraction.dir)) { dir.create(extraction.dir, recursive = T) }

watersheds <-
  get_wbd(
    template = map[, 'geometry'],
    label = county_name,
    extraction.dir = extraction.dir)
if('list' %in% class(watersheds)) { watersheds <- do.call(rbind, watersheds) }

ggplot(map, aes(geometry = geometry)) +
  theme_bw() + 
  geom_sf() + 
  geom_sf(data = watersheds, aes(geometry = geom), alpha = 0)
```

Lets download the actual streams.

**Hint**: `get_nhd()` has build in looping.
This means we can use the full `watersheds` to get all the streams in all the watersheds in the same call.
**However**, it is slow to the point of being unusable.
Strongly consider using a manual loop.
`interactive()` is used to hide the `txtProgressBar` when knitting.

**Hint**: The more layers on a chart, the harder it can be to tell them apart.
We add in an `fill = 'orange'` to help York county pop.

```{r}
hydro <- vector(mode = 'list', length = nrow(watersheds))
if(interactive()) { pb = txtProgressBar(min = 0, max = nrow(watersheds), style = 3) }
for(i in 1:nrow(watersheds)) {
  tmp <-
    get_nhd(
      template = watersheds[i,],
      label = sprintf('%s_%s', county_name, watersheds$huc12[i]),
      extraction.dir = extraction.dir)
  hydro[[i]] <- tmp$Flowline
  if(interactive()) { setTxtProgressBar(pb, i) }
}
if(interactive()) { close(pb) }
hydro <- do.call(rbind, hydro)

ggplot(map, aes(geometry = geometry)) +
  theme_bw() + 
  geom_sf(color = 'black', fill = 'orange') + 
  geom_sf(data = watersheds, aes(geometry = geom), alpha = 0) + 
  geom_sf(data = hydro, aes(geometry = geom), color = 'blue')
```
